---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ciccr

<!-- badges: start -->
<!-- badges: end -->

The goal of ciccr is to implement methods for carrying out causal inference in case-control studies.

## Installation

You can install the released version of ciccr from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ciccr")
```

Alternatively, you can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools") # uncomment this line if devtools is not installed yet 
devtools::install_github("sokbae/ciccr")
```
## Example

We first call the ciccr package.

```{r example}
library(ciccr)
```

To illustrate the usefulness of the package, we use the dataset ACS that is included in the package. This dataset is an extract from American Community Survey (ACS) 2018, restricted to white males residing in California with at least a bachelor's degree. The ACS is an ongoing annual survey by the US Census Bureau that provides key information about US population. We use the following variables:  

```{r}
  y = ACS$topincome
  t = ACS$baplus
  x = ACS$age
```

- The binary outcome `y` is defined to be one if a respondent's annual total pre-tax wage and salary income is top-coded. In the sample extract, the top-coded income bracket has median income $565,000 and the next highest income that is not top-coded is $327,000.   

- The binary treatment `t` is defined to be one if a respondent has a master's degree, a professional degree, or  a doctoral degree.

- The covariate `x` is age in years and is restricted to be between 25 and 70.

The original ACS survey is not from case-control sampling but we construct a case-control sample by the following procedure: 

1. The case sample is composed of 921 individuals whose income is top-coded.
2. The control sample of equal size is randomly drawn without replacement from the pool of individuals whose income is not top-coded. 

We now construct cubic b-spline terms with three inner knots using the age variable.

```{r}
  x = splines::bs(x, df = 6)
```
 
Using the retrospective sieve logistic regression model, we estimate the average of the log odds ratio conditional on the case sample by

```{r}
  results_case = avg_retro_logit(y, t, x, 'case')
  results_case$est
  results_case$se
```

Here, option `'case'` refers to conditioning on the event that income is top-coded. 

Similarly, we estimate the average of the log odds ratio conditional on the control sample by

```{r}
  results_control = avg_retro_logit(y, t, x, 'control')
  results_control$est
  results_control$se
```

Here, option `'control'` refers to conditioning on the event that income is not top-coded. 

We carry out causal inference by

```{r}
  cicc(y, t, x, 0.2)
```

Here, 0.2 is the specified upper bound for unknown case probability. If it is not specified, the default choice for `p` is `p = 1`.

```{r}
  cicc(y, t, x)
```

More to be added

# Reference

Sung Jae Jun and Sokbae Lee. Causal Inference in Case-Control Studies. 
https://arxiv.org/abs/2004.08318.

